---
title: "Untitled"
output: github_document
---
### Load packages
```{r load packages and default options, message = FALSE}
# Load packages
library(tidyverse)
library(broom)
library(knitr)
library(modelr)

# Set default figure options
knitr::opts_chunk$set(
  fig.width = 6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


## Abstract



## Introduction (brief context & background of the problem)

dataset of breast cancer patients from a prospective study. 

information was collected at baseline, survival months: length of following up, status: survival status

interested in predicting risk of death based on some/or all predictors.

Variables in dataset:

1. Age
2. Race
3. Marital Status
4. T Stage: Adjusted AJCC 6th T
5. N Stage: Adjusted AJCC 6th N
6. 6th Stage: Breast Adjusted AJCC 6th Stage
7. Differentiate
8. Grade
9. A Stage: Regional — A neoplasm that has extended; Distant — A neoplasm that has spread to
parts of the body remote from
10. Tumor Size: Each indicates exact size in millimeters.
11. Estrogen Status
12. Progesterone Status
13. Regional Node Examined
14. Reginol Node Positive
15. Survival Months
16. Status: Dead / Alive



```{r data import, echo = FALSE}

breast_cancer <- read_csv('data/Project_2_data.csv') |> 
  janitor::clean_names() |> 
  rename(regional_node_positive = reginol_node_positive) |> 
  mutate(
    race = as_factor(race),
    marital_status = as_factor(marital_status),
    x6th_stage = as_factor(x6th_stage),
    differentiate = as_factor(differentiate),
    a_stage = as_factor(a_stage),
    root_tumor_size = sqrt(tumor_size),
    estrogen_status = as_factor(estrogen_status),
    progesterone_status = as_factor(progesterone_status),
    status = case_match(status,
                        "Dead" ~ 1,
                        "Alive" ~ 0),
    regional_prop = regional_node_positive/regional_node_examined
  ) |> 
  select(-c(regional_node_positive, regional_node_examined, survival_months,
            t_stage, n_stage, grade, tumor_size))
```


```{r}
breast_cancer = breast_cancer %>%
  mutate(main_stage = case_when(
    str_detect(x6th_stage, "III") ~ "III",
    str_detect(x6th_stage, "II") ~ "II",
    str_detect(x6th_stage, "I") ~ "I",
  ),
  race2 = ifelse(race == "White", "White", "POC"))
```



```{r construct full model}
logistic_model =
  breast_cancer |> 
  glm(status ~ . + estrogen_status*progesterone_status, data = _, family = binomial())
```

```{r optimize model AIC}
opt_model =
  logistic_model |> 
  MASS::stepAIC(
    direction = "both",
    k = 2, trace = 0)
```

```{r}
opt_model |> 
  glance()

opt_model |> 
  tidy()

opt_model |> 
  car::vif()
```

Cooks distance 

```{r}
i = cooks.distance(opt_model)
plot(i,type="h")
```

Residuals 

```{r}
D = residuals(opt_model)
plot(D);abline(h=0,col="red")
```

Fitted

```{r}
p = predict(opt_model,type="response")

plot(x = p, y = D)
```


```{r}
y = breast_cancer$status
n = length(y)
yhat = rep(0,n)
yhat[p >= .5] = 1
```

```{r}
library(pROC)
tmp = roc(y,p)
plot(tmp)
print(tmp$auc)
```

```{r}
#White
library(pROC)
tmp = roc(y[breast_cancer$race == "White"],p[breast_cancer$race == "White"])
plot(tmp)
print(tmp$auc)
```

```{r}
#non White
library(pROC)
tmp = roc(y[breast_cancer$race != "White"],p[breast_cancer$race != "White"])
plot(tmp)
print(tmp$auc)
```

```{r}
#Black
library(pROC)
tmp = roc(y[breast_cancer$race == "Black"],p[breast_cancer$race == "Black"])
plot(tmp)
print(tmp$auc)
```

```{r}
#Other
library(pROC)
tmp = roc(y[breast_cancer$race == "Other"],p[breast_cancer$race == "Other"])
plot(tmp)
print(tmp$auc)
```


```{r}
D = (y - p)
plot(D);abline(h=0,col="red")
```

```{r}
data.frame(y = y, u = p, res_sq = (y - p)^2, race = breast_cancer$race,
           index = 1:4024) %>%
  ggplot(aes(x = index, y = res_sq, color = race)) + geom_point() +
  stat_smooth()
```




```{r add prediction and residual}
breast_cancer$probs = 
  
```

```{r qq plot}
breast_cancer |> 
  ggplot(aes(sample = resid)) +
  stat_qq() +
  stat_qq_line()
```

```{r resid plot}
breast_cancer |> 
  ggplot(aes(x = pred, y = resid)) +
  geom_point()
```




```{r}
breast_cancer %>%
  ggplot(aes(x = `x6th_stage`)) +
  geom_bar()
```

```{r}
breast_cancer %>%
  ggplot(aes(x = differentiate)) +
  geom_bar()
```

```{r}
## Commented out because grade is removed at data import
#breast_cancer %>%
  #ggplot(aes(x = grade)) +
  #geom_bar()
```

```{r}
breast_cancer %>%
  ggplot(aes(x = `estrogen_status`)) +
  geom_bar()
```

```{r}
breast_cancer %>%
  ggplot(aes(x = `race`)) +
  geom_bar()
```

```{r}
breast_cancer %>%
  ggplot(aes(x = status)) +
  geom_bar()
```

```{r tumor size transformation hist, echo = FALSE}
tumor_plot = 
  breast_cancer |> 
  mutate(tumor_size = root_tumor_size^2) |> 
  ggplot(aes(x = tumor_size)) + 
  geom_histogram(alpha = 0.8, col = I("black")) +
  labs(
    x = "Tumor size (mm)",
    y = "Frequency"
  )
  
tumor_plot_sq = 
  breast_cancer |> 
  ggplot(aes(x = root_tumor_size)) + 
  geom_histogram(alpha = 0.8, col = I("black")) +
  labs(
    x = "Square root of tumor size",
    y = "Frequency"
  )

ggpubr::ggarrange(tumor_plot, tumor_plot_sq)
```

